---
title: "Heatmaps in R"
author: "Kayvan Jalali & Carlos Puerta"
date: "9/27/2023"
output:
  beamer_presentation:
    keep_tex: no
    theme: CambridgeUS
    slide_level: 2
    incremental: no
fontsize: 10pt
classoption: compress
header-includes:
  \hypersetup{colorlinks,citecolor=orange,filecolor=red,linkcolor=brown,urlcolor=blue}
  \usepackage{amsmath,color,amsthm,amssymb,pdfpages,algorithm2e,dsfont}
  \newcommand{\vv}{\vspace{2mm}}
  \newcommand{\p}{\mathbb{P}}
  \newcommand{\e}{\mathbb{E}}
  \newcommand{\etr}{\overline{\text{err}}}
  \newcommand{\ete}{\text{Err}_{\mathcal{T}}}

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_chunk$set(dev = "pdf")

library(nycflights13)
library(ggplot2)
library(pheatmap)
library(tidyverse)
library(RColorBrewer)
```

```{r, echo=FALSE}

## ggplot dummy data
x = LETTERS[1:5]
y = paste0("var", seq(1, 5))
data = expand.grid(X = x, Y = y)
data$Z = runif(25, 0, 5)

## Pheat dummy data
test = matrix(rnorm(25), 20, 10)
test[1:10, seq(1, 10, 2)] = test[1:10, seq(1, 10, 2)] + 3
test[11:20, seq(2, 10, 2)] = test[11:20, seq(2, 10, 2)] + 2
test[15:20, seq(2, 10, 2)] = test[15:20, seq(2, 10, 2)] + 4
colnames(test) = paste("Test", 1:10, sep = "")
rownames(test) = paste("Gene", 1:20, sep = "")

```

\tableofcontents

# What is a heatmap?

A heatmap is graphic that will display your data in a colorful grid. This is great for seeing trends and patterns in your data.

Generally, heatmaps are used to represent data where you have 2 categorical variables with a third continuous variable, though this is not explicitly necessary.

### Advantages
\begin{itemize}
  \item Adds another variable
  \item Easier to interpret
  \item Color is easy to parse
\end{itemize}
 
### Disadvantages
\begin{itemize}
  \item Need right choice of colors and data 
  \item Only useful if clear 
  \item A different plot may be better
\end{itemize}

## Examples

```{r, echo=FALSE, fig.width = 2.3, fig.height = 2.3, warning=FALSE, message=FALSE}
ggplot(data, aes(X, Y, fill = Z)) +
  geom_tile() + coord_equal() + theme_minimal()

ggplot(mtcars, aes(x = wt, y = mpg, fill = disp)) +
  geom_point(shape = 21) + theme_minimal()
```

## When to use a heatmap

Heatmaps are applicable to both continuous and discrete variables. When working with discrete variables, distinguishing between similar colors can be challenging. To improve visibility, it is recommended to use a color palette of no more than nine colors, such as ROYGBIV, black, and white.

In the case of continuous variables, you have the option to use a sequential 
color scale, transitioning from a lighter hue to a darker one or vice versa. 
Alternatively, you can create a divergent color scale, transitioning between two distinct colors. A divergent color scale may be more useful when the center and ends
of the value range are meaning full. 

## ggplot2

You can create a simple and quick heatmap using the geom_tile() function in ggplot2, but these heatmaps are very limited in their functionality.

```{r, echo=TRUE, fig.width = 4, fig.height = 2, warning=FALSE, message=FALSE}
ggplot(data, aes(X, Y, fill = Z)) +
  geom_tile() + coord_equal()
```

# Pheatmap Package

Pheatmap (Pretty Heatmaps) is a package for R that supercharges heatmaps and allows you to create incredibly complex and helpful heatmaps. Although it is slightly more difficult to use when compared to ggplot, it is highly specialized and focused. It can plot more than 3 variables in a single heatmap, and more importantly, it groups rows and/or columns.

The grouping specifically can be incredibly helpful for visually spotting trends and similarities.

## Dendrogram

Dendrograms, create a higherarchy of similar groups. Eventhough they can be standalone, and are not always seen with heatmaps, pheatmap includes dendrograms by default. Groups that are most similar get pooled together until there are no more groupings possible. The distance between groups directly reflects the difference between groups. By default these groupings are made using the rows or columns euclidean distance.

### A Bad Tournament

You can think of it as a tournament bracket where seeding determines your matchups, your initial matches will be the strongest teams against each other and the weakest teams against each other. The distance between the connections symbolizes the overall difference in skill.

## Pheatmap example data creation 

## Basic Functions
```{r}

df <- diamonds %>%
  group_by(cut,clarity) %>%
  summarise(avg_carat =  mean(carat) ) %>%
  pivot_wider(names_from = clarity, values_from = avg_carat) %>% 
  column_to_rownames('cut')

```

## Pheatmap with no Arguments 

```{r}
pheatmap(df)
```

## Make Annotations

```{r}
# Annotations 
ann1 <- diamonds %>%
  group_by(cut) %>%
  summarise(avg_len =  mean(x) ) %>%
  column_to_rownames('clarity')

ann2 <- diamonds %>%
  group_by(cut) %>%
  summarise(avg_wid =  mean(y) ) %>%
  column_to_rownames('cut')
ann2


# Annotations colors
ann_colors = list(
  avg_len = c( "violet", "yellow"),
  avg_wid = c( brewer.pal(9,'YlGn'))
  )

```

## Lots of Function

```{r}

pheatmap(df,
         labels_row = paste(rownames(df),1:5),
         labels_col = paste(colnames(df), 1:8 ), # overlay 
         cluster_rows = TRUE,
         cluster_cols = TRUE,
         color = colorRampPalette(c("black","blue"))(100),
         border_color = 'red',
         #color = brewer.pal(9,'Blues'),
         #clustering_distance_rows = "correlation",
         clustering_distance_cols = "correlation",
         main = "Average Price by Cut and Clarity",
         show_rownames = TRUE,
         show_colnames = TRUE,
         annotation_col = ann1,
         annotation_row = ann2,
         annotation_colors = ann_colors,
         legend_breaks = c(.5,.9,1.3),
        # legend_labels = c(".5", ".9", "1.3","Average Price"),
         display_numbers = TRUE,
         number_color = 'green',
        fontsize = 6
         )

```

```{r}

# cont. example data
n <- 20
x1 = matrix(1:n, nrow = n)
x2 = matrix(1:n, nrow = 1)

mat = x1 %*% x2 # dim will be nxn

disc_mat = matrix(sample(1:4,n^2,replace=TRUE),nrow=n)



```

## Just Pheatmap
```{r, fig.width = 6, fig.height = 3}
pheatmap(mat)
```

## Dendrogram
```{r, fig.width = 6, fig.height = 3}
pheatmap(mat, cluster_rows = FALSE, cluster_cols = FALSE)
```

## Color
```{r, fig.width = 6, fig.height = 3}
# example why sequential palette is good
pheatmap(mat,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  color = colorRampPalette(c("white", "red"))(100),
  display_numbers = round(var(mat))
)
```

## Pheatmap example
```{r, fig.width = 6, fig.height = 3}
pheatmap(mat,
  color = colorRampPalette(c("white", "red"))(100),
  display_numbers = round(var(mat))
)
```

## Pheatmap example
```{r, fig.width = 6, fig.height = 3}
# discrete example
pheatmap(disc_mat,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  color = c("blue", "red", "yellow", 1)
)
```

## Pheatmap example
```{r, fig.width = 6, fig.height = 3}
pheatmap(disc_mat,
  color = c("blue", "red", "yellow", 1)
)
```


## More Complex Example

## Genes

\begin{figure}
  \centering
  \includegraphics[width=0.5\columnwidth]{figures/genes.png}
\end{figure}

## Code for Genes


```{r, echo=TRUE, warning=FALSE, message=FALSE, eval=FALSE}
topGenes = order(rowVars(Biobase::exprs(x)),
                 decreasing = TRUE)[1:500]
rowCenter = function(x) { x - rowMeans(x) }
pheatmap(rowCenter(dfx[topGenes, ]),
  show_rownames = FALSE,
  show_colnames = FALSE,
  breaks = seq(-5, +5, length = 101),
  annotation_col = pData(x)[, c("sampleGroup", "Embryonic.day",
                                "ScanDate", "genotype")],
  annotation_colors = list(
    sampleGroup = groupColor,
    genotype = c(`FGF4-KO` = "chocolate1", `WT` = "azure2"),
    Embryonic.day = setNames(brewer.pal(9, "Blues")[c(3, 6, 9)],
                             c("E3.25", "E3.5", "E4.5")),
    ScanDate = setNames(brewer.pal(nlevels(x$ScanDate), "YlGn"),
                        levels(x$ScanDate))
  )
)
```

# Applying Heatmaps

## MPG
Let's do it with a simple dataset we're familiar with.
```{r, echo=TRUE, warning=FALSE, message=FALSE}
mpg2 <- mpg %>%
  group_by(manufacturer, class) %>%
  mutate(avg_mpg = (cty + hwy) / 2) %>%
  summarize(avg_mpg = mean(avg_mpg)) %>%
  spread(class, avg_mpg) %>%
  column_to_rownames("manufacturer")
mpg2[is.na(mpg2)] = 0
mpg2[1:5, 1:6]
```

## MPG
```{r, echo=TRUE, warning=FALSE, message=FALSE, fig.width = 6, fig.height=3}
pheatmap(mpg2)
```

## MPG

That didn't look great. Let's clean it up.
```{r, echo=TRUE, eval=FALSE, warning=FALSE, message=FALSE}
pheatmap(mpg2,
  angle_col = 45,
  color = colorRampPalette(
    c("gray", "red", "orange", "yellow", "darkgreen")
  )(10),
)
```

## MPG
```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.width = 6, fig.height=3.7}
pheatmap(mpg2,
  angle_col = 45,
  color = colorRampPalette(
    c("gray", "red", "orange", "yellow", "darkgreen")
  )(10),
)
```

## NYC13Flights

We want to plot the total number of hours flights were late arrivals each month, broken up by each carrier.

```{r, echo=TRUE, warning=FALSE, message=FALSE}
flights2 <- flights %>%
  filter(arr_delay > 0) %>%
  group_by(month, carrier) %>%
  summarize(total_delay = sum(arr_delay / 60)) %>%
  spread(carrier, total_delay) %>%
  column_to_rownames("month")
```

## NYC13Flights
We can now plot it.

```{r, echo=TRUE, eval=TRUE, fig.width=10, fig.height=4.5, warning=FALSE, message=FALSE}
pheatmap(flights2)
```

## NYC13Flights

Additionally, we would like to overlay the total airtime (in days) for each month. We can do this like as well.

```{r, echo=TRUE, warning=FALSE, message=FALSE}
airtime <- flights %>%
  filter(arr_delay > 0) %>%
  group_by(month) %>%
  summarize(airtime = sum(air_time / (60 * 24))) %>%
  column_to_rownames("month")
```

```{r, echo=TRUE, eval=FALSE, fig.width=6, fig.height=4, warning=FALSE, message=FALSE}
pheatmap(flights2, annotation_row = airtime)
```

## NYC13Flights
```{r, echo=FALSE, fig.width=6, fig.height=4, warning=FALSE, message=FALSE}
pheatmap(flights2, annotation_row = airtime)
```

## NYC13Flights

We also know certain airlines are preferred, so we would like to see the total number of flights by airline.

```{r, echo=TRUE, warning=FALSE, message=FALSE}
flightcounts <- flights %>%
  group_by(carrier) %>%
  summarize(total_flights = n()) %>%
  column_to_rownames("carrier")

colnames(flightcounts) = "Total Flights"
```

```{r, echo=TRUE, eval=FALSE, fig.width=6, fig.height=4, warning=FALSE, message=FALSE}
pheatmap(flights2, annotation_row = airtime, annotation_col = flightcounts)
```

## NYC13Flights
```{r, eval=TRUE, fig.width=6, fig.height=4, warning=FALSE, message=FALSE}
pheatmap(flights2, annotation_row = airtime, annotation_col = flightcounts)
```

## NYC13Flights

If we're not interested in the dendrogram, we can remove it easily.
```{r, echo=TRUE, eval=FALSE, fig.width = 6, fig.height=4, warning=FALSE, message=FALSE}
pheatmap(flights2,
  annotation_row = airtime,
  annotation_col = flightcounts,
  cluster_row = FALSE,
  cluster_cols = FALSE
)
```

## NYC13Flights

```{r, fig.width = 6, fig.height=4, warning=FALSE, message=FALSE}
pheatmap(flights2,
  annotation_row = airtime,
  annotation_col = flightcounts,
  cluster_row = FALSE,
  cluster_cols = FALSE
)
```


## NYC13Flights

Finally, we can clean up our main legend, change the color scheme, and very clearly show missing values for our final figure.
```{r, echo=TRUE, eval=FALSE, fig.width = 4, fig.height=4, warning=FALSE, message=FALSE}
pheatmap(flights2,
  annotation_row = airtime,
  annotation_col = flightcounts,
  cluster_row = FALSE,
  cluster_cols = FALSE,
  legend_breaks = c(150, 1200, 2400),
  legend_labels = c("Low", "Medium", "High"),
  color = RColorBrewer::brewer.pal(9, "PuRd"),
  na_col = "black"
)
```

## NYC13Flights

```{r, fig.width = 10, fig.height=6, warning=FALSE, message=FALSE}
pheatmap(flights2,
  annotation_row = airtime,
  annotation_col = flightcounts,
  cluster_row = FALSE,
  cluster_cols = FALSE,
  legend_breaks = c(150, 1200, 2400),
  legend_labels = c("Low", "Medium", "High"),
  color = RColorBrewer::brewer.pal(9, "PuRd"),
  na_col = "black"
)
```

## You can also choose to go bananas

\begin{figure}
  \centering
  \includegraphics{figures/genes.png}
\end{figure}

# Wrapping up
